In this comparison we include some tables that would be excluded in the main analysis because they are not RCTs. These tables are still useful here though, as they allow us to test the accuracy of the algorithm on similar tables.


# from 3_compare_algorithm_hand.Rmd

The results are ordered by the reason for rejection from the algorithm.

```{r}
both = full_join(excluded_hand, excluded_algorithm, by='pmcid') %>%
  select(pmcid, hand, algorithm) %>%
  unique() %>%
  arrange(algorithm) # sort by reason
ftab = flextable(both) %>%
  fontsize(size=8, part = 'all') %>%
  autofit()
ftab
```


#### Table as above but just for first three rows of the table

```{r}
stats = filter(matches, row <= 3) %>%
  select(pmcid, statistic.y, starts_with('match')) %>%
  pivot_longer(cols=starts_with('match')) %>%
  filter(!is.na(value)) %>% # remove missing comparisons (both stats needed) %>%
  group_by(statistic.y, name) %>%
  summarise(n = n(), 
            match = sum(value)) %>%
  mutate(percent = 100*match/n,
         stat = case_when(
           name == 'match_stat1' ~ 'First',
           name == 'match_stat2' ~ 'Second'
         ),
         cell = paste(match, ' / ', n, ' (', roundz(percent,1),')', sep='')) %>%
  select(statistic.y, stat, cell) %>%
  rename('Statistic type' = 'statistic.y',
         'Statistic' = 'stat',
         'n / N (%)' = 'cell')
#
ftab = flextable(stats) %>%
  theme_box() %>%
  merge_v(j=1) %>%
  autofit()
ftab

```

#### From 5_summary_stats.Rmd

### Boxplot by table column

```{r}
# 
sizes = dplyr::select(table_data, pmcid, column, sample_size) %>%
  unique()
#
gplot = ggplot(data=sizes, aes(x=factor(column), y=sample_size))+
  geom_boxplot()+
  scale_y_log10()+
  ylab('Sample size')+
  xlab('Table column')+
  theme_bw()
gplot
```

### Parametric fit to log sample size

```{r, include=FALSE}
fg = fitdist(data=log(sizes$sample_size), distr="gamma") # spits stuff to screen
max_size = max(sizes$sample_size)
N = nrow(sizes)
pred = dgamma(x=log(1:max_size), shape=fg$estimate[1], rate=fg$estimate[2])
gamma_line = data.frame(row = log(1:max_size), fitted = pred*N)
# make random data for easy plotting in histogram
rgamma = data.frame(rgamma(n=N, shape=fg$estimate[1], rate=fg$estimate[2]))
names(rgamma) = 'sample_size'
rgamma = mutate(rgamma, sample_size = round(exp(sample_size)))
```

```{r}
# 
to_plot = bind_rows(sizes, rgamma, .id='source') %>%
  mutate(source = ifelse(source==1, 'Observed data', 'Fitted from gamma'))
#
gplot = ggplot(data=to_plot, aes(x=log(sample_size)))+
  geom_histogram(fill='skyblue') +
  ylab('Count')+
  xlab('Log-transformed sample size')+
  theme_bw()+
  facet_wrap(~source)
gplot
```